Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-26 manuscript.rmd
1. edited symbols. 1. mfac figures now shown by censuses

\begin{equation}
\Psi_\textrm{soil} = \Psi_e\left(\frac{\theta}{\theta_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively. $\mathrm{\Psi}_e$, $\theta_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.

@@

We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) $\cong$ $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical time domain reflectometry (TDR) probe over 0-15 cm depth ($\theta_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\theta_{\textrm{0-0.15 m}}$, $\theta_{\textrm{0.4 m}}$ and $\theta_{\textrm{1 m}}$ refer to plot-wide averages).

@@

\begin{equation}
\begin{split}
\overline{O} = \frac{1}{6}\sum \textrm{RMSE}^{*}\textrm{(ET,}E_{v} + E_{g}) + \textrm{RMSE}^{*}(q_{\textrm{run}},q_{\textrm{over}} + q_{\textrm{drai}}) + \textrm{RMSE}^{*}(\theta_{\textrm{0-0.15~m}}, {W_{\textrm{liq,}\overline{z}}}) \\
& + \textrm{RMSE}^{*}(\theta_{\textrm{0.15~m}}, W_{\textrm{liq,`r soil.depths[4]`m}}) + \textrm{RMSE}^{*}(\theta_{\textrm{0.4~m}}, W_{\textrm{liq,`r soil.depths[6]`~m}}) + \textrm{RMSE}^{*}(\theta_{1~m}, W_{\textrm{liq,`r soil.depths[8]`~m}})
(\#eq:obj-func)
\end{split}
\end{equation}

based on monthly sums for both ET and $q$~run~ and daily averages for $\theta$s. $W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$, and $Z$ = {`r paste(soil.depths[1:4], collapse=", ")`} m. Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.

\   Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}(\theta_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

@@

To estimate effective rooting depth, we tested alternate empirical models for species 5-yearly census growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Because pre-dawn $\mathrm{\Psi}$~leaf~ approaches $\mathrm{\Psi}$~soil~, we used the latter in leaf hydraulic vulnerability curves to predict maximum diurnal $K_\textrm{leaf}$, which in turn controls growth (see Table \@ref(tab:symbols) for trait symbols and their definitions). $K_\textrm{leaf}$ in our models thus incorporates the effects of rooting depth and rhizosphere water potential. This approach, however, does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion). VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020]. Apart from stomatal control, leaf-drop may further limit water-uptake and growth. We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. Due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

\begin{equation}
K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}\Psi_{\textrm{soil},i|z,h}}
(\#eq:k-soil)
\end{equation}
\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth. Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization *h* we retained only one depth that had the maximum likelihood, and within those depths retained the ones that had Pearson's $r$ >= 0.71 as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :

@@

\begin{equation}
K_{leaf,s} = A_{s}e^{-B_{s}\Psi_{leaf_{s}}} + \epsilon_{s}
(\#eq:vc-exp)
\end{equation}
where $A$ and $B$ are fitted species-specific parameters, $\epsilon_{s}$ an error term. These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation. For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies. We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits; namely, WSG, the wood specific gravity, and LMA, the leaf mass per unit area (@Wright:2010; see SI for details). We fitted polynomial equations described as,

@@

Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with mortality rates significantly decreased with ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig. \@ref(fig:erd-mort-census-evg)). When deciduous species were included, ERD explained lower but significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig. \@ref(fig:erd-mort-census)). Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig. \@ref(fig:swp-droughts)). Soil water availability was not modeled for 1985-1990. Analysis of a species' time spent under significant soil drought (proportion of days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds (Fig. \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)). Species with ERD deeper than 2 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa).

@@

Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins (Fig. \@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Fig. \@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)). Deeper ERD species on average had more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk. By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{soil,~z = ERD}$ crossed ERD $\mathrm{\Psi}_{crit}$ (Fig. \@ref(fig:erd-mort-census-evg); Fig. \@ref(fig:erd-time)). Our results are consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019].

@@

\
\   Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decoupled climate events and demographic outcomes. Future studies could better constrain the ERD model with high frequency growth data such as those from dendrometer bands. At high temporal resolution, however, the role of reversible dehydration in tree diameter change increases [@Chitra-Tarak.2016; @Chitra-Tarak.2015], but that may provide an avenue to include stem water storage  and dynamic rooting depths in ERD models. Three of our exploratory ERD models included LAI seasonality, but we did not select them as they worsened the fit with growth data for many species (Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation. Future studies may further develop models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].

@@

\
Figure \@ref(fig:erd-time): Modeled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--% days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was below $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~50,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy tree species on Barro Colorado Island. Each circle represents a species. Circles are colored and slightly jittered horizontally for clarity.

@@

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time}")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr.jpeg"))
```

@@

```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($Psi$~soil,z~) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines). Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
```

@@

```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over 1990-2015, where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are colored and slightly jittered horizontally for clarity.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mean_mfac vs. rdi.gr_evergreen.jpeg"))
```

>>>>

\begin{equation}
\mathrm{\theta}_\textrm{soil} = \Psi_e\left(\frac{\mathrm{\theta}}{\mathrm{\theta}_{s}}\right)^{-b}
(\#eq:soil-ret-curve)
\end{equation}
where $\mathrm{\Psi}_e$ is the air-entry matric potential, $\theta_{s}$ is the saturated volumetric water content, and $b$, an index for soil pore-size distribution, respectively. $\mathrm{\Psi}_e$, $\theta_{s}$ and $b$ that best-fit the data were 200 (mm H~2~O), 0.51 (cm^3^ cm^-3^) and 10 (unitless), respectively.

@@

We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) $\cong$ $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical time domain reflectometry (TDR) probe over 0-15 cm depth ($\mathrm{\theta}_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).

@@

\begin{equation}
\begin{split}
\overline{O} = \frac{1}{6}\sum \textrm{RMSE}^{*}\textrm{(ET,}E_{v} + E_{g}) + \textrm{RMSE}^{*}(q_{\textrm{run}},q_{\textrm{over}} + q_{\textrm{drai}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0-0.15~m}}, {W_{\textrm{liq,}\overline{z}}}) \\
& + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.15~m}}, W_{\textrm{liq,`r soil.depths[4]`m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{\textrm{0.4~m}}, W_{\textrm{liq,`r soil.depths[6]`~m}}) + \textrm{RMSE}^{*}(\mathrm{\theta}_{1~m}, W_{\textrm{liq,`r soil.depths[8]`~m}})
(\#eq:obj-func)
\end{split}
\end{equation}

based on monthly sums for both ET and $q$~run~ and daily averages for $\mathrm{\theta}$s. $W_{\textrm{liq,}\overline{z}}$ is the average of daily $W_{\textrm{liq,} z}$, where $z$ $\in$ $Z$, and $Z$ = {`r paste(soil.depths[1:4], collapse=", ")`} m. Superscript * indicates that RMSE is standardized between 0 and 1 across the 5000 simulations.

\   Preliminary analyses showed that minimizing the objective function $\overline{O}$, that is, simultaneously fitting the model to all states and fluxes did not capture the soil moisture dynamics at 1 m depth well, our key priority for model calibration. To ensure model fit at 1 m depth, we first chose 300 parameter ensemble members that minimized $\textrm{RMSE}^{*}({\mathrm{\theta}_{\textrm{1 m}}, W_{\textrm{liq,`r soil.depths[8]` m}})$ (< 0.2), and among those 300 ensemble members we chose 100 with the lowest values for the objective function $\overline{O}$. See Table \@ref(tab:param) for the ranges of best-fit values for different parameters.

@@

To estimate effective rooting depth, we tested alternate empirical models for species 5-yearly census growth (Eq. \@ref(eq:M-chosen) and those in SI), which use effective rooting depth as an input. Many factors limit plant growth, including soil moisture, VPD, leaf area seasonality and radiation [@Lawrence.2019]. Because pre-dawn $\mathrm{\Psi}$~leaf~ approaches $\mathrm{\Psi}$~soil~, we used the latter in leaf hydraulic vulnerability curves to predict maximum diurnal $K$~leaf~, which in turn controls growth (see Table \@ref(tab:symbols) for trait symbols and their definitions). $K_\textrm{leaf}$ in our models thus incorporates the effects of rooting depth and rhizosphere water potential. This approach, however, does not account for the buffering effect of stem water storage capacitance on $\mathrm{\Psi}$~leaf~, and may thus overestimate ERD (see Discussion). VPD affects growth non-linearly, with growth increasing with VPD up-to a VPD threshold, then decreasing as leaf pores (stomata) close reducing water-uptake [@Grossiord.2020]. Apart from stomatal control, leaf-drop may further limit water-uptake and growth. We therefore regressed species-specific growth against multiplicative or additive effects of one or more factors calculated daily and averaged over 5-yearly censuses: VPD, $K_\textrm{leaf}$ and the leaf area index (LAI), all standardized between 0 and 1. Based on our preliminary analyses, additional incorporation of radiation did not improve model fitting. Due to the coincidence of higher temperature, higher radiation, and lower humidity during the dry season at BCI, the impact of VPD may as well be a combined impact of not only temperature and humidity, but also radiation.

@@

\begin{equation}
K_{\textrm{leaf},s,i|z,h} = A_{s}e^{-B_{s}\mathrm{\Psi}_{\textrm{soil},i|z,h}}
(\#eq:k-soil)
\end{equation}
\    By varying $z$, we estimated species effective rooting depths (ERD), defined as the depth $z$ at which soil water dynamics ($\mathrm{\Psi}_{\textrm{soil,} z}$) best explained observed dynamics of species growth. Using Eq. \@ref(eq:M-chosen), for a species $s$, we obtained likelihood based on Akaike's Information Criterion (AIC) and Pearson's correlation coefficient $r$ for each soil depth $z$ for each of the 100 best-fit hydrological realizations. For each hydrological realization *h* we retained only one depth that had the maximum likelihood, and within those depths retained the ones that had Pearson's $r$ >= 0.71 as a valid set. For a species $s$, we defined effective rooting depth (ERD) as the median of these set of the retained depths, which can be described as :

@@

\begin{equation}
K_{\textrm{leaf},s}} = A_{s}e^{-B_{s}\mathrm{\Psi}_{\textrm{leaf}_{s}}} + \epsilon_{s}
(\#eq:vc-exp)
\end{equation}
where $A$ and $B$ are fitted species-specific parameters, $\epsilon_{s}$ an error term. These 21 species included `r sp.n.data` of the `r length(erd.sp)` species selected for ERD estimation. For the rest `r sp.n.model` species, we obtained vulnerability curves using trait-based proxies. We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits; namely, WSG, the wood specific gravity, and LMA, the leaf mass per unit area (@Wright:2010; see SI for details). We fitted polynomial equations described as,

@@

Among evergreen species, ERD explained `r m.evg.r2.90`, `r m.evg.r2.05` and `r m.evg.r2.10`% variation in mortality rates during census intervals 1985-1990, 2000-2005 and 2005-2010, respectively, such that species with mortality rates significantly decreased with ERD ($p$ = `r m.evg.p.90`, `r m.evg.p.05` and `r m.evg.p.10`, respectively; Fig. \@ref(fig:erd-mort-census-evg)). When deciduous species were included, ERD explained lower but significant proportion of variation in mortality in those intervals ($R$^2^ = `r m.r2[1]`, `r m.r2[2]` and `r m.r2[3]`% and $p$ = `r m.p.90`, `r m.p.05` and `r m.p.10`, respectively; Fig. \@ref(fig:erd-mort-census)). Census intervals 2000-2005 and 2005-2010 were also the periods during which extreme soil droughts distinctly occurred in the wet season, effectively extending the dry season into the wet season (Fig. \@ref(fig:swp-droughts)). Soil water availability was not modeled for 1985-1990. Analysis of a species' time spent under significant soil drought (proportion of days $\mathrm{\Psi}_{\textrm{soil},z=\textrm{ERD}}$ < $\mathrm{\Psi}$~crit~) showed that species with shallower ERD spent greater proportion of days below critical hydraulic thresholds and increasingly so over 1995-2000, 2000-2005 and 2005-2010 (Fig. \@ref(fig:erd-time); S\@ref(fig:erd-time-evg)). Species with ERD deeper than 2 m did not cross species $\mathrm{\Psi}$~crit~ for the entire 25-yr period for which soil water potential $\mathrm{\Psi}_{\textrm{soil},z}$ was modeled (1990-2015). Soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` MPa (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`; median, 95% CI) through all the dry seasons and drought periods, which is above ($\mathrm{\Psi}$~crit~) of the most sensitive tree species (-0.28 MPa).

@@

Our study brings attention to the need for assessing drought sensitivity in terms of species realized drought exposure by accounting for the actual hydrological drought and trees' deep water access, beyond the universal drought indices and trees' above-ground hydraulic vulnerability. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins (Fig. \@ref(fig:erd-hyd)), that is, traits that are commonly identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through periods of extreme droughts (Fig. \@ref(fig:erd-mort-census-evg), S\@ref(fig:erd-mort-census), \@ref(fig:swp-droughts)). Deeper ERD species on average had more hydraulically efficient, but risky, suite of traits optimized for an environment of more reliable and greater water source at depth that resulted in a lower mortality risk. By contrast shallow ERD species had inefficient, but safer, xylem likely optimized for high variability in $\mathrm{\Psi}_\textrm{soil}$ in the shallow soil layers, and resulting in higher variance in mortality rates across censuses as $\mathrm{\Psi}_{\textrm{soil},z\textrm{ = ERD}}$ crossed ERD $\mathrm{\Psi}_\textrm{crit}$ (Fig. \@ref(fig:erd-mort-census-evg); Fig. \@ref(fig:erd-time)). Our results are consistent with a few studies that included three or four species each with known rooting or water sourcing depths and assessed their hydraulic traits and mortality rates during extreme droughts [@Johnson:2018fad; @Nardini:2015; @Venturas.2016; also see @Brum.2019].

@@

\
\   Our inverse ERD model was parameterized on 5-year growth data, with five data points over a 25-yr period, which decoupled climate events and demographic outcomes. Future studies could better constrain the ERD model with high frequency growth data such as those from dendrometer bands. At high temporal resolution, however, the role of reversible dehydration in tree diameter change increases [@Chitra-Tarak.2016; @Chitra-Tarak.2015], but that may provide an avenue to include stem water storage  and dynamic rooting depths in ERD models. Three of our exploratory ERD models included LAI seasonality, but we did not select them as they worsened the fit with growth data for many species (Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and leaf hydraulic vulnerability curves may be adequate to explain inter-census differences in growth (via stomatal control), but also acknowledge that our estimates of seasonality of LAI is a tentative estimate combining leaf-fall data and the timing of leaf-gain backtracked from leaf lifetime, and omits inter-annual variation. Future studies may improve models of LAI seasonality and make use of data from new technologies such as drone based monitoring of LAI in species-rich tropical forests [@Park.2019].

@@

\
Figure \@ref(fig:erd-time): Modeled species Effective Rooting Depth (ERD; mean $\pm$ 1SE in m) versus time spent below critical hydraulic threshold--% days for which $\mathrm{\Psi}_{\textrm{soil},z}$ was below $\mathrm{\Psi}$~crit~, defined as $\mathrm{\Psi}$~50,leaf~, over each of the five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for `r length(erd.mrate.sp)` canopy tree species on Barro Colorado Island. Each circle represents a species. Circles are colored and slightly jittered horizontally for clarity.

@@

```{r erd-time, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time}")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr.jpeg"))
```

@@

```{r swp-all-depths, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-swp-all-depths}Daily mean soil water potential ($\\Psi_{\\textrm{soil},z}$) by depth $z$ (m; panels) over 1990-2018 predicted from each of the the 100 best-fit ensembles (colored lines). Vertical gray lines demarcate median dates of the demographic censuses at the BCI 50 ha plot."}
knitr::include_graphics(file.path("figures/2019-10-14_5000/best-fits/psi_model_daily_all_depths_params.top.few_full.jpeg"))
```

@@

```{r erd-time-evg, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "40%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-time-evg}Modelled species Effective Rooting Depth (ERD; mean $\\pm$ 1SE in m) versus time spent below critical hydraulic threshold--the number of days for which $\\mathrm{\\Psi}_{\\textrm{soil},z}$ was below $\\mathrm{\\Psi}$~crit~, defined as $\\mathrm{\\Psi}$~80,leaf~, over five census intervals (1990-2015), where $z$ is the soil depth matching species ERD--for ", nrow(mrate.depth.mean %>% subset(deciduous == "E")), " canopy species found on Barro Colorado Island. Each circle represents a species. Circles are colored and slightly jittered horizontally for clarity.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/mfac vs. rdi.gr_evergreen.jpeg"))
```
